`r glue::glue('inferCNV results for {params$library_id} with {params$reference_name} reference')`

### 2025-05-09 analyses/infercnv-consensus-cell-type/template-notebooks/SCPCP000015_explore-infercnv-results.Rmd

# Extract a version of the data frame with just the infercnv metadata, not CNVs themselves
infercnv_metadata_df <- infercnv_df |>
  dplyr::select(
    -starts_with("has_"),
    -starts_with("proportion_"),
    -starts_with("top_")
  )

@@

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  dplyr::select(
    barcodes,
    starts_with("has_cnv_")
  ) |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each
  dplyr::group_by(barcodes) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  dplyr::ungroup() |>
  # bring back metadata
  dplyr::inner_join(
    infercnv_metadata_df,
    by = "barcodes"
  ) |>
  # join in SCE information
  dplyr::left_join(celltype_df) |>
  dplyr::left_join(umap_df) |>
  # add cell type grouping column, with counts
  dplyr::mutate(
    celltype = dplyr::case_when(
      cell_group == "reference" ~ "reference cell",
      stringr::str_detect(ewing_annotation, "tumor") ~ "tumor cell",
      .default = "other query cell"
    )
  ) |>
  # add columns with cell type & group counts
  dplyr::add_count(celltype, name = "celltype_count") |>
  dplyr::add_count(cell_group, name = "cell_group_count") |>
  dplyr::mutate(
    celltype_n = glue::glue("{celltype} (n={celltype_count})"),
    cell_group_n = glue::glue("{cell_group} (n={cell_group_count})")
  )

# Set the celltype_n factor order based on label, not frequency
# We don't know the exact labels due to `(n=)`, so use grep
has_cnv_df$celltype_n <- as.factor(has_cnv_df$celltype_n)
present_values <- levels(has_cnv_df$celltype_n)
has_cnv_df$celltype_n <- forcats::fct_relevel(
  has_cnv_df$celltype_n,
  c(
    grep("reference", present_values, value = TRUE),
    grep("tumor", present_values, value = TRUE),
    grep("other", present_values, value = TRUE)
  )
)
```

@@

right_plot <- ggplot(has_cnv_df) +
  aes(y = forcats::fct_rev(celltype_n), x = total_cnv_per_cell, fill = celltype_n) +
  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
  labs(y = "Cell type (number of cells)") +
  theme(legend.position = "bottom")

@@

celltype_umap <- ggplot(library_has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = celltype_n) +
  geom_point(size = 0.2, alpha = 0.3) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(title = "Cell types") +
  umap_theme

@@

Now, we'll look at CNVs across chromosomes, specifically considering the `proportion_scaled_cnv_` output from the HMM.

@@

```{r}
# First, construct data frame with proportion_scaled_ instead of has_cnv
prop_cnv_df <- infercnv_df |>
  dplyr::select(
    library_id,
    barcodes,
    starts_with("proportion_scaled_")
  ) |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  # more tidying
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
  # remove extra chromosomes and cells not in this library
  dplyr::filter(
    !stringr::str_starts(chr, "chrMT"),
    !stringr::str_starts(chr, "chrGL"),
    library_id == params$library_id
  ) |>
  # set chr factor levels, with ones used for validation first
  dplyr::mutate(
    chr = stringr::str_remove(chr, "^chr"),
    chr = factor(chr, levels = 1:22),
    chr = forcats::fct_relevel(chr, as.character(validation_df$chromosome))
  ) |>
  # finally, join in the cell types
  dplyr::inner_join(
    has_cnv_df |>
      dplyr::select(barcodes, celltype_n)
  )

prop_cnv_df
```

@@

```{r fig.height=12, fig.width=14}
ggplot(loss_df) +
  aes(x = proportion, color = celltype_n) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 4 # this way the first row is the validation ones only
  ) +
  ggtitle("Proportion of CNV loss across chromosomes for each cell type group") +
  theme(strip.text = element_text(size = 8))
```

```{r fig.height=12, fig.width=14}
ggplot(gain_df) +
  aes(x = proportion, color = celltype_n) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 4 # this way the first row is the validation ones only
  ) +
  ggtitle("Proportion of CNV gain across chromosomes for each cell type group") +
  theme(strip.text = element_text(size = 8))
```

>>>>

# Extract a version of the data frame with just the infercnv metadata, not CNVs themselves
# Note that we keep this separate for circumstances when the metadata has to be joined back in with results
infercnv_metadata_df <- infercnv_df |>
  dplyr::select(
    -starts_with("has_"),
    -starts_with("proportion_"),
    -starts_with("top_")
  )

@@

```{r}
# First, construct data frame of total `has_cnv` counts, with other variables needed for plots
has_cnv_df <- infercnv_df |>
  tidyr::pivot_longer(
    starts_with("has_cnv_"),
    names_to = "chr",
    values_to = "cnv"
  ) |>
  # sum all values for each
  dplyr::group_by(barcodes) |>
  dplyr::summarize(total_cnv_per_cell = sum(cnv)) |>
  dplyr::ungroup() |>
  # bring back metadata
  dplyr::inner_join(
    infercnv_metadata_df,
    by = "barcodes"
  ) |>
  # join in SCE information
  dplyr::left_join(celltype_df) |>
  dplyr::left_join(umap_df) |>
  # add cell type grouping column, with counts
  dplyr::mutate(
    celltype = dplyr::case_when(
      cell_group == "reference" ~ "reference cell",
      stringr::str_detect(ewing_annotation, "tumor") ~ "tumor cell",
      .default = "other query cell"
    )
  ) |>
  # add columns with cell type & group counts
  dplyr::add_count(celltype, name = "celltype_count") |>
  dplyr::add_count(cell_group, name = "cell_group_count") |>
  dplyr::mutate(
    celltype = forcats::fct_relevel(celltype, c("reference cell", "tumor cell", "other query cell")),
    # set celltype labels with (n=)
    celltype = forcats::fct_relabel(
      celltype,
      # https://stackoverflow.com/questions/70510771/how-can-i-use-fct-relabel-but-maintain-original-factor-levels-or-how-can-i-set
      ~ glue::glue("{levels(celltype)} (n = {celltype_count[match(levels(celltype), celltype)]})")
    ),
    cell_group_n = glue::glue("{cell_group} (n={cell_group_count})")
  )
```

@@

right_plot <- ggplot(has_cnv_df) +
  aes(y = forcats::fct_rev(celltype), x = total_cnv_per_cell, fill = celltype) +
  ggridges::geom_density_ridges2(alpha = 0.7, scale = 0.85) +
  labs(y = "Cell type (number of cells)") +
  theme(legend.position = "bottom")

@@

celltype_umap <- ggplot(library_has_cnv_df) +
  aes(x = UMAP1, y = UMAP2, color = celltype) +
  geom_point(size = 0.2, alpha = 0.3) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  labs(title = "Cell types") +
  umap_theme

@@

Now, we'll look at CNVs across chromosomes, specifically considering the `proportion_scaled_cnv_` output from the HMM focusing on gain and loss events.

@@

```{r}
# First, construct data frame with proportion_scaled_ instead of has_cnv
prop_cnv_df <- infercnv_df |>
  tidyr::pivot_longer(
    starts_with("proportion_scaled_"),
    names_to = "cnv_type_raw",
    values_to = "proportion"
  ) |>
  # more tidying
  dplyr::mutate(cnv_type_raw = stringr::str_remove(cnv_type_raw, "^proportion_scaled_")) |>
  tidyr::separate(cnv_type_raw, sep = "_", into = c("cnv_type", "chr")) |>
  # remove extra chromosomes and cells not in this library
  dplyr::filter(
    !stringr::str_starts(chr, "chrMT"),
    !stringr::str_starts(chr, "chrGL"),
    library_id == params$library_id
  ) |>
  # set chr factor levels, with ones used for validation first
  dplyr::mutate(
    chr = stringr::str_remove(chr, "^chr"),
    chr = factor(chr, levels = 1:22),
    chr = forcats::fct_relevel(chr, as.character(validation_df$chromosome))
  ) |>
  # finally, join in the cell types
  dplyr::inner_join(
    has_cnv_df |>
      dplyr::select(barcodes, celltype)
  )
```

@@

```{r fig.height=12, fig.width=14}
ggplot(loss_df) +
  aes(x = proportion, color = celltype) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 4 # this way the first row is the validation ones only
  ) +
  ggtitle("Proportion of CNV loss across chromosomes for each cell type group") +
  theme(strip.text = element_text(size = 8))
```

```{r fig.height=12, fig.width=14}
ggplot(gain_df) +
  aes(x = proportion, color = celltype) +
  geom_density() +
  facet_wrap(
    vars(chr),
    scales = "free",
    ncol = 4 # this way the first row is the validation ones only
  ) +
  ggtitle("Proportion of CNV gain across chromosomes for each cell type group") +
  theme(strip.text = element_text(size = 8))
```
